Data Exploratory Analyses/Plots

### 2022-12-28 DataExploratoryAnalyses.Rmd
Working on exploratory analyses

```{r,echo=FALSE,results='hide',message=FALSE,warning=FALSE}
allPrecipData <- read.csv(paste0(dataPath,"NEON_PrecipitationDataDailydata_gapFilled.csv"))
allPrecipData$year <- lubridate::year(allPrecipData$date)
allPrecipData <- allPrecipData %>% group_by(year,siteID) %>% mutate(cumPrecip=cumsum(precipBulk)) #Calculates the cummulative sum of precip for each site-year for each date

@@

@@

#For each PFT-phenophase combination: Does the timing and Standard Deviation in DOYs differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

@@

#For each PFT-phenophase combination: Does the length and Standard Deviation differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

@@

@@

  print(ggplot(winterLengthDat,
         aes(winter_sd,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
    geom_density_ridges2(scale=1)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=90))+
    xlab("SD")+
    ylab("Site ID (Ordered by Latitude)")+
    facet_wrap(~growthForm)+
    ggtitle(paste("SD in Winter Length Per Individual: Colored-",NEON_phenophase_names[initialP]))+
    scale_x_continuous(limits=c(0,100)))
}
# #Based on Leaf Numbers
# leavesDat <- read.csv(file=paste0(dataPath,'NEON_PhenologyObservations/NEON_PhenoObservationData_',gsub(" ","",NEON_phenophase_names[p]),'.csv'))
# initialGrowth <- leavesDat %>% filter(phenophaseIntensity=="5-")

@@

@@

```{r}

>>>>

#Load Covariate Data
```{r,echo=FALSE,results='hide',message=FALSE,warning=FALSE}
allPrecipData <- read.csv(paste0(dataPath,"NEON_PrecipitationDataDailydata_gapFilled.csv"))
allPrecipData$year <- lubridate::year(allPrecipData$date)
allPrecipData <- allPrecipData %>% group_by(year,siteID) %>% mutate(cumPrecip=cumsum(precipBulk)) #Calculates the cummulative sum of precip for each site-year for each date

@@

What I learned: Measured phenophases differ a lot by PFT. Evergreen don't have an end of growing season measurement. Many just have leaves as their changing phenophase.

@@

What I learned: There aren't large sample sizes per individual and the sample sizes range between 1 and 8.

#For each PFT-phenophase combination: Does the timing in DOYs differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

```{r,echo=FALSE,results='hide',message=FALSE,warning=FALSE,fig.width=10,fig.height=20}
for(p in seq_along(NEON_phenophase_names)){
  subDat <- read.csv(file=paste0(dataPath,'NEON_PhenologyObservations/NEON_PhenoObservationData_',gsub(" ","",NEON_phenophase_names[p]),'.csv'))
  subDat <- subDat %>% filter(phenophaseIntensity == mediumIntensity_phenophases[p])
  subDat <- subDat[,c('siteID','individualID','Mycorrhizal.type','decimalLatitude','dayOfYear','growthForm')]
  subDat <- subDat %>% filter(growthForm %in% c("Deciduous broadleaf","Drought deciduous broadleaf","Forb","Graminoid"))

  if(nrow(subDat)>2){
    print(ggplot(subDat,aes(dayOfYear,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
            geom_density_ridges2(scale=1)+
            theme_bw()+
            theme(axis.text.x = element_text(angle=90))+
            xlab("DOY")+
            ylab("Site ID (Ordered by Latitude)")+
            facet_wrap(~growthForm)+
            ggtitle(NEON_phenophase_names[p]))
          #+
           # scale_x_continuous(limits=c(0,100)))
  }
}

What I learned: For most sites, AM budburst earlier and senescened earlier than ECM

#For each PFT-phenophase combination: Does the Standard Deviation in DOYs differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

@@

What I learned: EcM seem maybe have less sd in timing in budburst. SD in Colored leaves doesn't seem to differ as much between mycorrhizae types.

New question: Do they break buds earlier when corrected for site or latitude effects?

#For each PFT-phenophase combination: Does the length of winter and summer differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

@@

  winterLengthDat <- inner_join(coloredLeavesDat,budBurstDat,
                            by=c("winterYear","individualID",'siteID',"Mycorrhizal.type","decimalLatitude",
                                     "growthForm")) %>% filter(growthForm %in% c("Deciduous broadleaf","Drought deciduous broadleaf","Evergreen broadleaf","Evergreen conifer","Forb","Graminoid","Semi-evergreen broadleaf"))%>% 
    mutate(winterLength=as.Date(date.y)-as.Date(date.x))

  summerLengthDat <- inner_join(coloredLeavesDat,budBurstDat,
                                by=c("year","individualID",'siteID',"Mycorrhizal.type",
                                     "decimalLatitude","growthForm")) %>% filter(growthForm %in% c("Deciduous broadleaf","Drought deciduous broadleaf","Evergreen broadleaf","Evergreen conifer","Forb","Graminoid","Semi-evergreen broadleaf")) %>%
    mutate(summerLength=as.Date(date.x)-as.Date(date.y))

  print(ggplot(winterLengthDat,
         aes(winterLength,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
    geom_density_ridges2(scale=1)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=90))+
    xlab("Winter Length (days)")+
    ylab("Site ID (Ordered by Latitude)")+
    facet_wrap(~growthForm)+
    ggtitle(paste("Winter Length Per Individual: Colored-",NEON_phenophase_names[initialP])))

    print(ggplot(summerLengthDat,
         aes(summerLength,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
    geom_density_ridges2(scale=1)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=90))+
    xlab("Summer Length (days)")+
    ylab("Site ID (Ordered by Latitude)")+
    facet_wrap(~growthForm)+
    ggtitle(paste("Summer Length Per Individual: Colored-",NEON_phenophase_names[initialP])))

}
# #Based on Leaf Numbers
# leavesPhenophases <- determinePhenophaseTimingForLeaves(dataPath) %>% 
#   filter(growthForm %in% c("Deciduous broadleaf","Drought deciduous broadleaf","Evergreen broadleaf","Evergreen conifer","Forb","Graminoid","Semi-evergreen broadleaf")) 
# 
# greenupDat <- leavesPhenophases %>% dplyr::select(-c(greendownDate,greendownDOY)) %>%
#   mutate(winterYear=(year),year=as.numeric(year))
# greendownDat <- leavesPhenophases %>% dplyr::select(-c(greenupDate,greenupDOY)) %>%
#   mutate(winterYear=(year+1),year=as.numeric(year))
# 
# winterLengthDat <- inner_join(greendownDat,greenupDat,
#                                 by=c("winterYear","individualID",'siteID',"Mycorrhizal.type",
#                                      "decimalLatitude","growthForm")) %>%
#   mutate(winterLength=as.Date(greenupDate)-as.Date(greendownDate)) %>%
#     group_by(individualID) %>% mutate(winter_sd=sd(winterLength,na.rm=TRUE)) %>%
#     ungroup() %>%
#     dplyr::select(individualID,winter_sd,siteID,Mycorrhizal.type,decimalLatitude,growthForm) %>% unique()
#   
#   summerLengthDat <- inner_join(greendownDat,greenupDat,
#                                 by=c("year","individualID",'siteID',"Mycorrhizal.type",
#                                      "decimalLatitude","growthForm")) %>%
#   mutate(summerLength=as.Date(greendownDate)-as.Date(greenupDate)) %>%
#     group_by(individualID) %>% mutate(summer_sd=sd(summerLength,na.rm=TRUE)) %>%
#     ungroup() %>%
#     dplyr::select(individualID,summer_sd,siteID,Mycorrhizal.type,decimalLatitude,growthForm) %>% unique()
#   
#   ggplot(winterLengthDat,
#          aes(winter_sd,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
#     geom_density_ridges2(scale=1)+
#     theme_bw()+
#     theme(axis.text.x = element_text(angle=90))+
#     xlab("SD")+
#     ylab("Site ID (Ordered by Latitude)")+
#     facet_wrap(~growthForm)+
#     ggtitle(paste("SD in Winter Length Per Individual: Number of Leaves"))+
#     scale_x_continuous(limits=c(0,50))
#   
#     ggplot(summerLengthDat,
#          aes(summer_sd,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
#     geom_density_ridges2(scale=1)+
#     theme_bw()+
#     theme(axis.text.x = element_text(angle=90))+
#     xlab("SD")+
#     ylab("Site ID (Ordered by Latitude)")+
#     facet_wrap(~growthForm)+
#     ggtitle(paste("SD in Summer Length Per Individual: Number of Leaves"))+
#     scale_x_continuous(limits=c(0,50))
#

What I learned: EcM maybe has a slightly shorter winter and AM maybe has a slightly shorter summer.

#For each PFT-phenophase combination: Does the SD of winter and summer lengths differ between phenophases differ between EcM and AM associated plants acrross all sites and within each site?

@@

  print(ggplot(winterLengthDat,
         aes(winter_sd,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
    geom_density_ridges2(scale=1)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=90))+
    xlab("SD")+
    ylab("Site ID (Ordered by Latitude)")+
    facet_wrap(~growthForm)+
    ggtitle(paste("SD in Winter Length Per Individual: Colored-",NEON_phenophase_names[initialP]))+
    scale_x_continuous(limits=c(0,100)))

    print(ggplot(summerLengthDat,
         aes(summer_sd,fct_reorder(siteID,decimalLatitude),fill=Mycorrhizal.type)) +
    geom_density_ridges2(scale=1)+
    theme_bw()+
    theme(axis.text.x = element_text(angle=90))+
    xlab("SD")+
    ylab("Site ID (Ordered by Latitude)")+
    facet_wrap(~growthForm)+
    ggtitle(paste("SD in Summer Length Per Individual: Colored-",NEON_phenophase_names[initialP]))+
    scale_x_continuous(limits=c(0,100)))
}

@@

What I learned: The standard deviations in winter and summer lengths don't really seem to differ between EcM and AM.

@@

```{r}
p <- 2 #Bud burst 
  print(NEON_phenophase_names[p])
  subDat <- read.csv(file=paste0(dataPath,'NEON_PhenologyObservations/NEON_PhenoObservationData_',gsub(" ","",NEON_phenophase_names[p]),'.csv'))
  subDat <- subDat %>% filter(phenophaseIntensity == mediumIntensity_phenophases[p],
                              growthForm=="Deciduous broadleaf")
  subDat$year <- as.numeric(subDat$year)

  allIndividuals <- unique(subDat$individualID)
  subDat <- subDat %>% dplyr::select(siteID,date,dayOfYear,individualID,canopyPosition,
                           height,growthForm,Mycorrhizal.type,year)

  allOutput <- matrix(nrow=0,ncol=10)
  colnames(allOutput) <- c('siteID','date','dayOfYear','individualID',
                           'canopyPosition','height','growthForm',
                           'Mycorrhizal.type','year','phenophaseOccurred')
  for(i in seq_along(allIndividuals)){
    if(i %% 100==0){
      print(i)
    }
    subSubDat <- subDat %>% filter(individualID==allIndividuals[i])
    for(j in 1:nrow(subSubDat)){
      currentRow <- subSubDat[j,]
      newOutput <- currentRow %>% slice(rep(1:n(), each = (subSubDat$dayOfYear[j]-1)))
      newOutput <- as.data.frame(newOutput)
      newOutput$date <- seq((as.Date(currentRow$date)-currentRow$dayOfYear+1),
                            (as.Date(currentRow$date)-1),by="day")
      newOutput$dayOfYear <- as.integer(row.names(newOutput))
      newOutput$phenophaseOccurred <- FALSE
      currentRow$phenophaseOccurred <- TRUE
      newOutput <- rbind(newOutput,currentRow)
      newOutput <- newOutput[(max((nrow(newOutput)-29),0)):nrow(newOutput),]

      allOutput <- rbind(allOutput,newOutput)
    }

  }
  OGallOutput <- allOutput
  allOutput$date <- as.character(as.Date(allOutput$date))
  allOutput <- left_join(allOutput,allPrecipData[,-3],by=c('siteID','date','year'))
  allOutput <- left_join(allOutput,allTempData,by=c('siteID','date','year'))
  allOutput <- left_join(allOutput,allSunlightTimes,by=c('siteID','date'))
  allOutput <- allOutput %>% filter(year!=2022)
  allOutput$Mycorrhizal.type <- as.factor(allOutput$Mycorrhizal.type)

  #Deciduous Broadleaf
  subDatDB <- allOutput %>% filter(growthForm=="Deciduous broadleaf",canopyPosition %in% c("Mostly shaded","Partially shaded","Full sun","Full shade"))

  # mdl_gam1 <- gam(dayOfYear ~ s(GDD_15,k=10) + s(GDD_5,k=10) + s(GDD_0,k=10) + s(GDD_10,k=10),data=subDatDB,
  #                  method="REML",
  #                  na.action="na.fail")

  mdl_gam <- gam(phenophaseOccurred ~ s(GDD_10) + s(daylength),data=subDatDB,
                   method="REML",
                   na.action="na.fail")

  mdl <- lm <- lm(phenophaseOccurred ~ (GDD_10) + (daylength),data=subDatDB)
  #######********Editing here: How to get it to predict a categorical response variable?****

  gam.check(mdl_gam,page=1)
  plot.gam(mdl_gam,residuals=TRUE,pch=20,cex=5,shade=TRUE,page=1,all.terms = TRUE)
  gam.check(mdl_gam1,page=1)

  plot(subDatDB$daylength,subDatDB$dayOfYear,pch=20)
  plot(subDatDB$GDD_5,subDatDB$dayOfYear,pch=20)
  plot(subDatDB$cumPrecip,subDatDB$dayOfYear,pch=20)

  mdl_gam2 <- gam(dayOfYear ~ s(daylength,by=Mycorrhizal.type) + s(GDD_10,by=Mycorrhizal.type) + s(cumPrecip,by=Mycorrhizal.type) + ,data=subDatDB,
                   method="REML",
                   na.action="na.fail")

    plot.gam(mdl_gam,residuals=TRUE,pch=20,cex=5,shade=TRUE,page=1,all.terms = TRUE)
      gam.check(mdl_gam1,page=1)
